% -*- TeX-engine: luatex; fill-column: 72; eval: (auto-fill-mode -1); eval: (visual-fill-column-mode 1); eval: (visual-line-mode 1); eval: (adaptive-wrap-prefix-mode 1) -*-
\documentclass{article}
%% You will need the "Turing Technical Notes" package.
%% See the (very short) installation instructions here: https://github.com/alan-turing-institute/ttn
%%
\usepackage{ttn}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage[Ragged, size=footnote, shape=up, alerton]{sidenotesplus}
\usepackage[dvipsnames]{xcolor}
\usepackage{siunitx}
%% 
%%\usepackage[style=authoryear]{biblatex}
%%\addbibresource{notes.bib}
%%
%% \defn takes an optional star and an optional argument.
%% If the optional argument is present, it is used for the marginal note; if the argument is missing, the default is the mandatory argument. If the command is starred, no marginal note is made. 
\NewDocumentCommand\defn{sO{#3}m}{\textit{#3}\IfBooleanF{#1}{\sidenote*{\textcolor{MidnightBlue}{#2}}}}
%%
\title{Mathematics for language models}
\author{The Language Models Reading Group}
\date{\today}
%%
\begin{document}%\maketitle
%%
\newcommand{\vocab}{\ensuremath{\mathcal{V}}}
\newcommand{\emptytext}{\ensuremath{\varepsilon}}
\newcommand{\lang}{\ensuremath{\mathcal{L}}}
\maketitle

\section{Languages}

Fix, once and for all, a finite set,~\vocab, called the \defn{vocabulary}. The elements of \vocab\ will be called~\defn{words}. Typically, \vocab\ is in fact the set of words in some lexicon but it might also be a set of tokens, or a set of letters.  We leave open the question of whether \vocab\ includes punctuation, capitalised words, and so on.  The notation $\left|\vocab\right|$ usually refers to the number of elements of~\vocab.

We are usually interested in sequences of words. A finite sequence of words is called a~\defn{text}.  For example, if $w_1$, $w_2$, \dots, $w_n$ are words, then the sequence $(w_1, w_2, \dotsc, w_n)$ is a text.  The zero-length sequence consisting of no words is also a text: we denote it by~\defn[empty text, \emptytext]{\emptytext}. 

We usually make use of a notational abbreviation for texts in which the symbols representing the words are simply adjoined; for example: $w_1 w_2 \dotsb w_n$.  But if we are writing a specific text, consisting of actual words, then we surround the text in single quotes and we may visually separate the words.  \sidenote*{What we have called ``texts'' are often called ``strings,'' or ``sentences.''  But our texts will include many (actual) sentences; and to a programmer, a string is a sequence of letters.}

\emph{Example}: Let the vocabulary be the set $\vocab=\{0, 1\}$.  Then \emptytext, `0', `1', `01', `1101', and `0101010' are texts.

\emph{Example}: Let the vocabulary be the set $\vocab = \{\text{the}, \text{on}, \text{cat}, \text{mat}, \text{sat}\}$.  Then `the', `mat mat mat', and `cat on cat on mat' are texts.

The set consisting of \emph{all} texts is denoted~$\vocab^*$. The `$*$' is called the \emph{Kleene star}.\sidenote{After Stephen Cole Kleene, 1909--1994. Kleene invented regular expressions in 1951, according to Wikipedia, ``to describe McCulloch-Pitts neural networks.'' Plus ça change!} It's the same $*$ as used in regular expressions---which Kleene invented---where it means ``zero or more repetitions of the preceding set of letters.'' It has essentially the same meaning here. Unless \vocab\ is empty, there are countably infinitely many elements of~$\vocab^*$.

A \defn{language}, $\lang$, is a subset of~$\vocab^*$.

\emph{Example}: The empty set, consisting of no texts, is a language, since $\emptyset\subset\vocab^*$. $\vocab^*$~itself is a language, since $\vocab^*\subset\vocab^*$.

\emph{Example}: Let $\vocab=\{0,1\}$ as before. A natural number, written in binary, is a text over this vocabulary. Consider the set of all natural numbers, expressed in binary; that is, $\lang = \{\text{0}, \text{1}, \text{10}, \text{11}, \text{100}, \dotsc\}$. This \lang\ is a language. (Note that \lang\ is a proper subset of $\vocab^*$ since, for example, it does not contain any texts starting with `0' apart from `0'.)

\emph{Example}: Let $\vocab=\{0,1\}$. A rational number between~0 and~1, may be written as a binary ``decimal.'' For example, the fraction $1/2$ is $0.1$ in binary. Consider the set of all such binary expressions, omitting the initial `$0.$'. We do \emph{not} obtain a language: this set contains infinite sequences (such as the binary expansion of $1/10$) and infinite sequences are not texts.

\emph{Example}: Let $\vocab =  \{\text{the}, \text{on}, \text{cat}, \text{mat}, \text{sat}\}$. The set consisting of the texts, `the cat sat on the mat', `the cat sat on the cat', `the mat sat on the cat', and `the mat sat on the mat' is a language.  

\emph{Exercise}: Let $\vocab=\emptyset$ be empty. How many languages over $\vocab$ are there? (Hint: first work out what $\vocab^*$ is.)

Sometimes we are interested only in sequences of words of a certain length. We denote by $\vocab^N$ the set of all sequences of $N$ words, for $N$ some natural number. (The notation is essentially that for the Cartesian product.)

\section{Surprise}

Suppose an event~$E$ occurs, one to which beforehand we had assigned a probability $P(E)$, with $0 < P(E) \le 1$.  How \emph{surprised} should we be to see it happen?  Two demands pin the answer down.  First, a more probable event ought to be less surprising: surprise should decrease as $P(E)$ increases, and a sure thing, $P(E)=1$, should cause no surprise at all.  Second, learning that two \emph{independent} events both occurred should surprise us by the \emph{sum} of the two separate surprises---even though their joint probability is the \emph{product} of the two separate probabilities.  A function that turns products into sums is a logarithm, and, up to the choice of base, there is essentially only one function meeting both demands.

We therefore define the \defn[Shannon information content]{surprise} of an event~$E$ to be
\[
  \operatorname{surprise}(E) = \log_2 \frac{1}{P(E)} = -\log_2 P(E).
\]
Because we have taken the logarithm to base~2, surprise is measured in \defn{bits}.\sidenote{Other bases are used too: base~$e$ gives \emph{nats}, and base~10 gives \emph{bans}, or \emph{hartleys}. Changing the base merely rescales surprise by a constant factor.}

The formula has exactly the shape we asked for.  When $P(E)=1$ the surprise is $\log_2 1 = 0$; as $P(E)$ shrinks towards~0 the surprise grows without bound; and for independent events $E$ and $F$, where $P(E\cap F)=P(E)\,P(F)$,
\[
  \operatorname{surprise}(E\cap F) = \operatorname{surprise}(E) + \operatorname{surprise}(F),
\]
since the logarithm of a product is the sum of the logarithms.

\emph{Example}: A fair coin lands heads with probability $1/2$.  The event of seeing heads therefore carries $\log_2 2 = 1$ bit of surprise---which is what makes the bit the natural unit.  Two tosses of the coin are independent, so the event `HH' has probability $1/4$ and surprise $\log_2 4 = 2$ bits: one bit for each toss, just as additivity promises.

\section{Entropy}

Fix a probability distribution~$P$ on the vocabulary~\vocab: that is, a function assigning to each word~$w$ a probability $P(w)\ge0$, with $\sum_{w\in\vocab}P(w)=1$.\sidenote{This is a distribution over single words, not over texts; and because \vocab\ is finite, all the sums in this section are finite.}  Now draw a word at random according to~$P$.  Before we see which word it is, how much surprise should we \emph{expect}?  Each particular word~$w$, regarded as the event that it is the one drawn, carries surprise $\operatorname{surprise}(w)=\log_2\tfrac{1}{P(w)}$; and the word~$w$ turns up a fraction $P(w)$ of the time; so the average surprise is found by weighting each word's surprise by its probability and summing.

This average is the \defn{entropy} of the distribution:
\[
  H(P) = \sum_{w\in\vocab} P(w)\operatorname{surprise}(w) = \sum_{w\in\vocab} P(w)\log_2\frac{1}{P(w)} = -\sum_{w\in\vocab} P(w)\log_2 P(w).
\]
Like surprise, entropy is measured in bits.  A word with $P(w)=0$ never occurs and contributes nothing to the average; we accordingly read the term $P(w)\log_2\tfrac{1}{P(w)}$ as~0 whenever $P(w)=0$.\sidenote{This is the right convention, since $p\log_2\tfrac1p \to 0$ as $p\to0^+$.}

Two extremes bound the entropy.  If the distribution is concentrated on a single word---some $w_0$ has $P(w_0)=1$ and every other word probability~0---then the outcome is never in doubt, and $H(P)=0$: no surprise, on average or otherwise.  At the other extreme, the \defn{uniform distribution}, $P(w)=1/\left|\vocab\right|$ for every word, is the most uncertain, and carries the largest entropy any distribution on \vocab\ can,\sidenote{That the uniform distribution maximises entropy is not obvious from the formula; it follows from the concavity of the logarithm.}
\[
  H(P) = \sum_{w\in\vocab}\frac{1}{\left|\vocab\right|}\log_2 \left|\vocab\right| = \log_2 \left|\vocab\right|.
\]
Every distribution on \vocab\ therefore has entropy between these, $0 \le H(P) \le \log_2\left|\vocab\right|$.

\emph{Example}: Let $\vocab=\{0,1\}$ and let $P$ be the fair coin, $P(0)=P(1)=1/2$.  Its entropy is $\tfrac12\log_2 2+\tfrac12\log_2 2 = 1$ bit---the $\log_2\left|\vocab\right|=\log_2 2$ promised for the uniform distribution on two words.

\emph{Example}: Let $\vocab=\{0,1\}$ and let $P$ be a biased coin, $P(0)=0.9$ and $P(1)=0.1$.  Then
\[
  H(P) = -0.9\log_2 0.9 - 0.1\log_2 0.1 \approx 0.47 \text{ bits},
\]
less than the~1 bit of the fair coin: a predictable coin is, on average, less surprising.

\section{Perplexity}

Entropy measures uncertainty in bits, on a logarithmic scale.  It is often handier to undo the logarithm and speak instead of an equivalent \emph{number of choices}.  The \defn{perplexity} of a distribution~$P$ on \vocab\ is the entropy, exponentiated:
\[
  \operatorname{perplexity}(P) = 2^{H(P)}.
\]
Because entropy was an arithmetic average of the surprises $\log_2\tfrac1{P(w)}$, exponentiating turns it into a \emph{geometric} average of the quantities $1/P(w)$ themselves:
\[
  \operatorname{perplexity}(P) = 2^{\sum_{w\in\vocab} P(w)\log_2\frac{1}{P(w)}} = \prod_{w\in\vocab}\left(\frac{1}{P(w)}\right)^{P(w)}.
\]

The value has a clean reading: perplexity is the \emph{effective number of equally-likely choices}.  The uniform distribution on \vocab\ makes this exact---its entropy is $\log_2\left|\vocab\right|$, so its perplexity is $2^{\log_2\left|\vocab\right|}=\left|\vocab\right|$, one choice for each word, just as it should be.  A distribution more predictable than uniform behaves like a uniform choice among \emph{fewer} words.

\emph{Example}: The fair coin has entropy~1 bit and perplexity $2^1 = 2$: two genuine choices.  The biased coin with $P(0)=0.9$ has entropy $\approx0.47$ bits and perplexity $2^{0.47}\approx1.38$: barely more than one real choice, since one outcome is nearly certain.

We can now read the paper's use of the word.  In practice we do not know the true distribution over words; instead we have a \emph{model} that, at each point in a text, assigns a probability to the next word, and a corpus of real text on which to judge it.  Write $\hat P(w_t \mid w_1^{t-1})$ for the probability the model assigns to the $t$-th word once it has seen the words before it.\sidenote{The hat marks a model's estimate rather than a true probability, and $\hat P(w_t\mid w_1^{t-1})$ is a \emph{conditional} probability: the model's probability for $w_t$ \emph{given} the preceding words $w_1,\dotsc,w_{t-1}$.  We will treat conditional probability, and this ``next word given the previous ones'' picture of language, in a later section.}  Each observed word then carries a surprise $\log_2\tfrac{1}{\hat P(w_t\mid w_1^{t-1})}$ under the model.  Averaging that surprise over all $N$ words of the corpus and exponentiating gives the model's perplexity \emph{on that corpus}:
\[
  2^{\frac1N\sum_{t=1}^N \log_2\frac{1}{\hat P(w_t\mid w_1^{t-1})}}
  \;=\;
  \left(\prod_{t=1}^N \frac{1}{\hat P(w_t\mid w_1^{t-1})}\right)^{\!1/N}.
\]
The phrase \emph{negative log-likelihood} unpacks in three steps.  The \defn{likelihood} of the corpus under the model is the probability the model assigns to the very words that occurred---the product $\prod_{t=1}^N \hat P(w_t\mid w_1^{t-1})$.  Its logarithm, the \defn{log-likelihood} $\sum_{t=1}^N \log_2 \hat P(w_t\mid w_1^{t-1})$, trades that product for a sum.  And because every probability is at most~1, each $\log_2 \hat P$ is at most~0, so the log-likelihood is negative (or zero); negating it yields a non-negative quantity to report.  The point to notice is that the negative log-likelihood of a single word is nothing new:
\[
  -\log_2 \hat P(w_t\mid w_1^{t-1}) = \log_2\frac{1}{\hat P(w_t\mid w_1^{t-1})} = \operatorname{surprise}(w_t),
\]
precisely its surprise under the model.  So the average negative log-likelihood is exactly the average surprise---the exponent in the perplexity formula above.

This is why the paper's two descriptions agree.  The quantity is Bengio et al.'s ``geometric average of $1/\hat P(w_t\mid w_1^{t-1})$ \dots\ also \dots\ the exponential of the average negative log-likelihood'': the right-hand side of the display is the geometric average, and the exponent on the left is that average negative log-likelihood.  A \emph{lower} perplexity means the model was, on average, less surprised by the words that actually occurred, and so predicted the text better.


\end{document}
